The ANthropological Notation Ontology (ANNO): A core ontology for annotating human bones and deriving phenotypes

### 2024-02-15 paper.tex
fix overful hboxes, remove todonotes package, add bibtex entry and citation

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
%\usepackage[T1]{fontenc}
\usepackage{todonotes}
\usepackage{cleveref}
\usepackage{csquotes}
\usepackage{aurl}
\usepackage{booktabs}
\usepackage{tabulary}
\usepackage{siunitx}
\usepackage{tablefootnote}
%\usepackage{dcolumn}
%\usepackage{placeins} %provides \FloatBarrier

\newcommand{\aw}{AnthroWorks3D}
\newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
\newcommand{\gfo}[1]{\href{https://www.onto-med.de/ontologies/gfo/#1}{\texttt{#1}}}
\newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
\newcommand{\latin}[1]{\emph{#1}}
\daurl{anno}{https://annosaxfdm.de/ontology/}
\daurl{gfo}{https://www.onto-med.de/ontologies/gfo\#}
\daurl{fma}{http://purl.org/sig/ont/fma/fma}
\daurl{ta2}{https://ta2viewer.openanatomy.org/?id=}

@@

\begin{document}

@@

\begin{itemize}
\item Specific bones can be found as a subclass of \emph{bone organ} (\aurl{fma}{224804}) but, among others, also in terms of tissue as \emph{bone tissue} (\aurl{fma}{224804}) with then varying terms such as \emph{bone tissue of hip bone} (\aurl{fma}{42854}) or \emph{bone of sacrum} (\aurl{fma}{43624}) and placed on the same level as histological structures like \emph{lamellar bone} (\aurl{fma}{224806}).
\item Different anatomical landmarks are classified taxonomically in various ways. Many landmarks are described as a \emph{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g. \emph{flat bone} (\aurl{fma}{7476}). However, the categorization of bone shape is not consistently defined across anatomical sources ~\citep{ anatomiedesmenschen}.
\todo{konrad bitte hier die neue Quelle zu Quelle 35 hinzunehmen}
To avoid these dependencies and potential incongruities, ANNO sidesteps taxonomic categorizations. Through location-based mapping, ANNO allows for the coexistence of diverse taxonomies.
\item On the other hand, another landmark, the \emph{external acoustic aperture} (\aurl{fma}{61301}), is classified as a subclass of \emph{orifice of skull} (\aurl{fma}{53133}), which falls under a different subtree of \emph{anatomical spaces} (\aurl{fma}{5897}) in FMA.
\item Osteometric landmarks are mixed with anatomical landmarks in a hierarchy (subclasses of \emph{anatomical point} (\aurl{fma}{9658}) that lacks coherence and detail.
\item The \emph{orbit} (\aurl{fma}{53074}), comprised of multiple landmarks, is classified without detailed segmentation into its constituent landmarks.
\end{itemize}

@@

\begin{table}[b]
\centering
\caption{Statistical information. Column values do not add up because some entities are defined in both subontologies.}
\label{tab:stats}
\begin{tabulary}{\textwidth}{lSSSSSSSSl}
\toprule
% Expressivity according to Protege: ANNOdc and combined: ALCHIQ(D), ANNOds: ALE
Ontology	&\textnormal{Classes}	&\textnormal{Object prop.}	&\textnormal{Data p.}	&\textnormal{Individuals}	&\textnormal{Annotation p.}	&\textnormal{Axioms}	&\textnormal{FMA links}				&\textnormal{TA links} &Expressivity\\
\midrule
ANNOdc		&41						&11								&1								&0							&19				&178					&11							&0					&OWL DL\\
ANNOds		&1166					&0								&0								&803						&43				&18231					&503						&206				&OWL DL\\
\midrule
combined	&1195					&11								&1								&803						&53				&18382					&514						&206				&OWL DL\\
\bottomrule
\end{tabulary}
\end{table}

>>>>

% add. options: [seceqn,secthm,crcready]
\documentclass[sw]{iosart2x}
%\usepackage[T1]{fontenc}
\usepackage{cleveref}
\usepackage{csquotes}
\usepackage{aurl}
\usepackage{booktabs}
\usepackage{tabulary}
\usepackage{siunitx}
\usepackage{tablefootnote}
%\usepackage{dcolumn}
%\usepackage{placeins} %provides \FloatBarrier
\newcommand{\aw}{AnthroWorks3D}
\newcommand{\anno}[1]{\href{https://annosaxfdm.de/ontology/#1}{\texttt{#1}}}
\newcommand{\gfo}[1]{\href{https://www.onto-med.de/ontologies/gfo/#1}{\texttt{#1}}}
\newcommand{\fma}[1]{\href{http://purl.org/sig/ont/fma/fma#1}{\texttt{#1}}}
\newcommand{\latin}[1]{\emph{#1}}
\daurl{anno}{https://annosaxfdm.de/ontology/}
\daurl{gfo}{https://www.onto-med.de/ontologies/gfo\#}
\daurl{fma}{http://purl.org/sig/ont/fma/fma}
\daurl{ta2}{https://ta2viewer.openanatomy.org/?id=}

@@

%\overfullrule=10pt % for debugging overfull hboxes
\begin{document}

@@

\begin{itemize}
\item Specific bones can be found as a subclass of \emph{bone organ} (\aurl{fma}{224804}) but, among others, also in terms of tissue as \emph{bone tissue} (\aurl{fma}{224804}) with then varying terms such as \emph{bone tissue of hip bone} (\aurl{fma}{42854}) or \emph{bone of sacrum} (\aurl{fma}{43624}) and placed on the same level as histological structures like \emph{lamellar bone} (\aurl{fma}{224806}).
\item Different anatomical landmarks are classified taxonomically in various ways.
Many landmarks are described as a \emph{zone of bone organ} (\aurl{fma}{10483}) with superclasses determined by bone shape), e.g. \emph{flat bone} (\aurl{fma}{7476}).
However, the categorization of bone shape is not consistently defined across anatomical sources ~\citep{anatomiedesmenschen,forensicmanual}.
To avoid these dependencies and potential incongruities, ANNO sidesteps taxonomic categorizations.
Through location-based mapping, ANNO allows for the coexistence of diverse taxonomies.
\item On the other hand, another landmark, the \emph{external acoustic aperture} (\aurl{fma}{61301}), is classified as a subclass of \emph{orifice of skull} (\aurl{fma}{53133}), which falls under a different subtree of \emph{anatomical spaces} (\aurl{fma}{5897}) in FMA.
\item Osteometric landmarks are mixed with anatomical landmarks in a hierarchy (subclasses of \emph{anatomical point} (\aurl{fma}{9658}) that lacks coherence and detail.
\item The \emph{orbit} (\aurl{fma}{53074}), comprised of multiple landmarks, is classified without detailed segmentation into its constituent landmarks.
\end{itemize}

@@

\begin{table}[b]
\centering
\caption{Statistical information. Column values do not add up because some entities are defined in both subontologies.}
\label{tab:stats}
\begin{tabular}{lS[table-format=4.0]SSS[table-format=3.0]SS[table-format=5.0]S[table-format=3.0]S[table-format=3.0]l}
\toprule
% Expressivity according to Protege: ANNOdc and combined: ALCHIQ(D), ANNOds: ALE
Ontology	&\textnormal{Classes}	&\textnormal{Object prop.}	&\textnormal{Data p.}	&\textnormal{Individuals}	&\textnormal{Annotation p.}	&\textnormal{Axioms}	&\textnormal{FMA links}				&\textnormal{TA links} &Expressivity\\
\midrule
ANNOdc		&41						&11								&1								&0							&19				&178					&11							&0					&OWL DL\\
ANNOds		&1166					&0								&0								&803						&43				&18231					&503						&206				&OWL DL\\
\midrule
combined	&1195					&11								&1								&803						&53				&18382					&514						&206				&OWL DL\\
\bottomrule
\end{tabular}
\end{table}
